---
title: "Notes on R"
author: "Ahmed Eltawil"
date: "2024-10-29"
output: 
  html_document: 
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Special numbers

- *Inf*: represents infinity = 1 / 0, it can also be used as 1 / inf = 0.
         you can also have -inf.
- *NaN*: represents undefined value "Not a number"; e.g. 0 / 0; can also be
         thought as a missing value.


# Data Types

R has 5 atomic classes:

- character
- numeric ( real numbers)
- integer
- complex
- logical ( True / False )

The most basic object is a *Vector*
- A vector can only contain objects of the same class
- Exception is List, which is a vector that can contain objects of different
  classes
- Empty vector can be created by calling *vector()*

## Vectors

### Creating a vector

#### 1- Using the *c()* function
```{r, echo = TRUE}
x1 <- c(0.5, 0.6)       ## numeric
x2 <- c(TRUE, FALSE)    ## logical
x3 <- c(T, F)           ## logical
x4 <- c("a", "b", "c")  ## character
x5 <- 9:29              ## integer
x6 <- c(1+0i, 2+4i)     ## complex
```
#### 2- Using the *vector()* function
```{r}
x_v <- vector("numeric", 10)
x_v
```

### Mixing Objects
When mixing different classes in one object, all must convert into one common
class
```{r}
y <- c(1.7, "a")    ## character
y <- c(TRUE, 2)     ## numeric
y <- c("a", TRUE)   ## character
```

### Explicit Coercion
Object can be explicitly coerced from one type to another using the **as. \***
functions if available
Example:
```{r}
x <- 0:6
class(x)
```
```{r}
class(as.numeric(x))
```
```{r}
as.logical(x)
```
```{r}
as.character(x)
```

On the other side, non-sensical coercion results in **NAs**

```{r}
x <- c("a", "b", "c")
as.numeric(x)
```
```{r}
as.logical(x)
```
```{r}
as.complex(x)
```

## Lists

A special type of vector that can contain elements of different classes

```{r}
x <- list(1, "a", TRUE, 1+4i)
x
```

## Matrices

A matrix is a vector with a *dimenstion* attribute, which is itself an integer
vector of length 2 **(nrow, ncol)**
```{r}
m <- matrix(nrow= 2, ncol= 3)
m
```
```{r}
dim(m)
```
```{r}
attributes(m) ## m has only one attribute which is dim of size 2
```
Matrices are constructed column-wise, so entries can be thought of starting
in the "upper left" corner and running down the columns.
```{r}
m <- matrix(data=1:6, nrow= 2, ncol= 3)
m
```
Matrices can be created directly also from vectors by adding *dim* attribute
```{r}
m <- 1:10
m
```
```{r}
dim(m) <- c(2, 5)
m
```
- Cbind-ing vs Rbind-ing
```{r}
x <- 1:3
y <- 10:12
cbind(x, y)
```
```{r}
rbind(x,y)
```

## Factors

Factors are used to represent categorial data.

- Treated specially by modeling functions *lm()* & *glm()*

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"))
x
```
Note that it has a special attribute which is called *Levels*
- Levels output the available labels within the factor starting from the
baseline level and upwards, and if the order or the levels are not given
explicitly by the user, the R engine will arrange them alphabetically.

The function table gives a frequency count of each label:
```{r}
table(x)
```
R engine assign an integers in an ascending order starting from the base level.
The *unclass()* function return the corresponding Level number to each label:
```{r}
unclass(x)
```
The order of the Levels can be set using *Levels* argument to *factor()*.
```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"), levels = c("yes", "no"))
x
```
Not that the order of the Levels has been reversed!

## Missing values

Denoted by:

- *NAN*: For missing mathematical operation
- *NA* : For everything else
- *NA* values have a class also, so there are integers NA, character NA, etc.
- A *NAN* value is also *NA* but the converse is not true.

Example:
```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)
```
```{r}
is.nan(x)
```
```{r}
x<- c(1, 2, NaN, NA, 4)
is.na(x)
```
```{r}
is.nan(x)
```

## Data Frames

Used to discribe tabular data

- Special type of list, where each element must have the same length
- Each element can be thought of as a column and the length of each element
  of this list is the number of rows.
- Unlike matrices, data frames can store different classes of objects in each
  column.
- Have special attribute called: *row.names*, most of the time it is just: 1, 2,
  3, ....
- Usually created by calling *read.table()* or *read.csv()*
- Can be converted to a matrix by calling *data.matrix()* but classes will be
  unified (resulted classes will be unexpected)
  
  Example:
```{r}
x <- data.frame(foo= 1: 4, bar= c(T, T, F, F))
x
```
```{r}
nrow(x)
```
```{r}
ncol(x)
```

## Names

R objects can also have names, which is very helpful for writing readable code
and self-describing objects.

```{r}
x <- 1:3
names(x)
```
```{r}
names(x) <- c("foo", "bar", "norf")
x
```
Lists can also have names
```{r}
x <- list(a=1, b=2, c=3)
x
```
For matrices, names are basicall, the rows and columns names:
```{r}
m <- matrix(data=1:4, nrow=2, ncol=2)
dimnames(m) <- list(c('a', 'b'), c('c', 'd'))
m
```

# Reading data

Most used APIs for reading data:
read.table
read.csv, same as read.table but in this case the separator is '*,*' by default.

## Estimating needed memory

First, to know if my Hardware is capable of laoding the data, do the following
calculation for a rough estimation on the needed memory:
$(nrow*ncol*8)*2$
Example: if we have a table consist of 1.500,000 rows and 120 columns, then
memory needed is equal to 2.68 GB

## Reading data optimization

Reading large data can be a huge overload on the processor.
To optimize the procedure, try the following:

- Read the help page for read.table, which contains many hints.
- Make a rough calculation for the needed memory to be stored.
- Set *comment.char*= "", if there is no comment lines in your file.
- Use the *colClasses* argument to specify the class for each column. Can make
it run much faster, often twice as fast. If all columns are numeric, then you
can just set *colClasses* = "numeric"

# Writing Data

Here we have two functions that write out data in textual form: dumping
and dputing. It's advantage over a table or csv file, is that it can
preserve the metadata (sacrificing some readability), so that another user
doesn't have to specify it all over again.

## Dputing R object

Can be used on a single R object
```{r}
y <- data.frame(a= 1, b= "a")
dput(y)
```
```{r}
dput(y, file= "y.R")
```

```{r}
new.y <- dget("y.R")
new.y
```

## Dumping R objects

Can be used on multiple R objects
```{r}
x <- "foo"
y <- data.frame(a= 1, b= "a")
dump(c("x", "y"), file="data.R")
rm(x,y)
```
Read the output again:
```{r}
source("data.R")
y
```
```{r}
x
```

# Connections

R can connect to several sources in order to read data

- file
- gzfile (gzip) (.gz)
- bzfile (bzip2) (.bz2)
- url

## structure of *file* function

```{r}
str(file)
```
1- description is the file name
2- open is a code indicating:
  - "r" -> read only
  - "w" -> write and/or create new file
  - "a" -> appending

## Connecting to a file

In case of reading the whole file with *read.csv* or *read.table*, no need
to open a connection as it is being done automatically.

But if you need to read specific lines of the file, a connection would make
sense in this case:
```{r}
## Code is commented as words file doesn't exist
## con <- gzfile("words.gz")
## x <- readLines(con, 10)
```

```{r}
con <- url("http://www.jhsph.edu", "r")
x<- readLines(con, 10)
x
```

# Subsetting

## Subsetting a vector

### Extracting one element from a list
```{r}
x <- c("a", "b", "c", "c", "d", "a")
print(x[1])
print(x[1:4])
print(x[x > "a"])
print(u <- x > "a")
print(x[u])
```

## Subsetting Lists

```{r}
x <- list(foo= 1:4, bar= 0.6)
x[1]
```
```{r}
x[[1]]
```
```{r}
x$foo ## $ is here same as [[]]
```
```{r}
x$bar
```
```{r}
x[["bar"]] ## same as x$bar
```
```{r}
x["bar"] ## single bracket always returns a list!
```

The operator *[[* can be used with computed indices; while *$* can only be
used with literal names.
```{r}
x <- list(foo= 1:4, bar= 0.6, baz= "hello")
name <- "foo"
x[[name]] ## computed index for 'foo'
```
```{r}
x$name ## element 'name' doesn't exist!
```

### Subsetting nested elements in a list

The *[[* can take an integer sequence
```{r}
x <- list(a= list(10, 12, 14), b= c(3.14, 2.81))
x[[c(1, 3)]]
```
```{r}
x[[1]][[3]] ## same as x[[c(1,3)]]
```
```{r}
x[[c(2,1)]]
```


### Extracting multiple elements from a list

```{r}
x <- list(foo= 1:4, bar= 0.6, baz= "hello")
x[c(1, 3)] ## accessing the first and third elements
```
**Note** *$* cannot be used to access multiple elements!

### Susetting a matrix

```{r}
x <- matrix(1:6, 2, 3)
print(x[1,2])
print(x[2,1])

print(x[1,])
print(x[,2])
```
By default, when a single element of the matrix is retrieved, it is returned as
a vector of length 1 rather than 1x1 matrix. this behavior cant be turned off
by setting ***drop = FALSE***
```{r}
x[1, 2, drop=FALSE]
```
```{r}
x[1, , drop=FALSE]
```

### Partial matching
```{r}
x <- list(aardvard= 1:5)
x$a
```
```{r}
x[["a"]] ## can't find element with exact name of "a"
```
```{r}
x[["a", exact= FALSE]] ## Get element with most similar name to "a"
```

### Removing NA values

#### is.na
```{r}
x <- c(1, 2, NA, 4, NA, 6)
bad <- is.na(x)
x[!bad]
```

#### complete.cases
*complete.cases* can be used with multiple vectors or data frames and return
only true on indices that are non NAs for all the vectors
```{r}
y <- c("a", "b", NA, "D", NA, "F")
good <- complete.cases(x, y)
good
```
```{r}
z <- c(NA, "b", "c", "D", NA, "F")
good2 <- complete.cases(y, z)
good2
```

#### Removing NA values in data frame
To be done.....👈

# Vectorized matrix operations

Many operations in R are vectorized making code more efficient, concise and
easier to read

```{r}
x <- 1:4; y <- 6:9
print(x+y)
print(x>2)
print(x>=2)
print(y==8)
print(x*y)
print(x/y)
```
```{r}
x <- matrix(1:4, 2, 2); y <- matrix(rep(10, 4), 2, 2)
print(x*y)
print(x/y)
```
In order to do a real matrix multiplication, the operator *%\*%* must be used
```{r}
x %*% y ## True matrix multiplication
```

**TODO: archive outside this file**
```{r}
inputData = read.csv("hw1_data.csv")
condInputData <- inputData[ inputData$Ozone > 31 & inputData$Temp > 90, ]
mean(condInputData$Solar.R, na.rm=TRUE)
condInputData <- inputData[inputData$Month == 6, ]
mean(condInputData$Temp, na.rm= TRUE)
condInputData <- inputData[inputData$Month == 5, ]
max(condInputData$Ozone, na.rm = TRUE)
x <- c(3, 5, 1, 10, 12, 6)
x[x%in%1:5] <- 0
x
```

