---
title: "Notes on R"
author: "Ahmed Eltawil"
date: "2024-10-29"
output: 
  html_document: 
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Special numbers

- *Inf*: represents infinity = 1 / 0, it can also be used as 1 / inf = 0.
         you can also have -inf.
- *NaN*: represents undefined value "Not a number"; e.g. 0 / 0; can also be
         thought as a missing value.


# Data Types

R has 5 atomic classes:

- character
- numeric ( real numbers)
- integer
- complex
- logical ( True / False )

The most basic object is a *Vector*
- A vector can only contain objects of the same class
- Exception is List, which is a vector that can contain objects of different
  classes
- Empty vector can be created by calling *vector()*

## Vectors

### Creating a vector

#### 1- Using the *c()* function
```{r, echo = TRUE}
x1 <- c(0.5, 0.6)       ## numeric
x2 <- c(TRUE, FALSE)    ## logical
x3 <- c(T, F)           ## logical
x4 <- c("a", "b", "c")  ## character
x5 <- 9:29              ## integer
x6 <- c(1+0i, 2+4i)     ## complex
```
#### 2- Using the *vector()* function
```{r}
x_v <- vector("numeric", 10)
x_v
```

### Mixing Objects
When mixing different classes in one object, all must convert into one common
class
```{r}
y <- c(1.7, "a")    ## character
y <- c(TRUE, 2)     ## numeric
y <- c("a", TRUE)   ## character
```

### Explicit Coercion
Object can be explicitly coerced from one type to another using the **as. \***
functions if available
Example:
```{r}
x <- 0:6
class(x)
```
```{r}
class(as.numeric(x))
```
```{r}
as.logical(x)
```
```{r}
as.character(x)
```

On the other side, non-sensical coercion results in **NAs**

```{r}
x <- c("a", "b", "c")
as.numeric(x)
```
```{r}
as.logical(x)
```
```{r}
as.complex(x)
```

## Lists

A special type of vector that can contain elements of different classes

```{r}
x <- list(1, "a", TRUE, 1+4i)
x
```

## Matrices

A matrix is a vector with a *dimenstion* attribute, which is itself an integer
vector of length 2 **(nrow, ncol)**
```{r}
m <- matrix(nrow= 2, ncol= 3)
m
```
```{r}
dim(m)
```
```{r}
attributes(m) ## m has only one attribute which is dim of size 2
```
Matrices are constructed column-wise, so entries can be thought of starting
in the "upper left" corner and running down the columns.
```{r}
m <- matrix(data=1:6, nrow= 2, ncol= 3)
m
```
Matrices can be created directly also from vectors by adding *dim* attribute
```{r}
m <- 1:10
m
```
```{r}
dim(m) <- c(2, 5)
m
```
- Cbind-ing vs Rbind-ing
```{r}
x <- 1:3
y <- 10:12
cbind(x, y)
```
```{r}
rbind(x,y)
```

## Factors

Factors are used to represent categorial data.

- Treated specially by modeling functions *lm()* & *glm()*

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"))
x
```
Note that it has a special attribute which is called *Levels*
- Levels output the available labels within the factor starting from the
baseline level and upwards, and if the order or the levels are not given
explicitly by the user, the R engine will arrange them alphabetically.

The function table gives a frequency count of each label:
```{r}
table(x)
```
R engine assign an integers in an ascending order starting from the base level.
The *unclass()* function return the corresponding Level number to each label:
```{r}
unclass(x)
```
The order of the Levels can be set using *Levels* argument to *factor()*.
```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"), levels = c("yes", "no"))
x
```
Not that the order of the Levels has been reversed!

## Missing values

Denoted by:

- *NAN*: For missing mathematical operation
- *NA* : For everything else
- *NA* values have a class also, so there are integers NA, character NA, etc.
- A *NAN* value is also *NA* but the converse is not true.

Example:
```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)
```
```{r}
is.nan(x)
```
```{r}
x<- c(1, 2, NaN, NA, 4)
is.na(x)
```
```{r}
is.nan(x)
```

## Data Frames

Used to discribe tabular data

- Special type of list, where each element must have the same length
- Each element can be thought of as a column and the length of each element
  of this list is the number of rows.
- Unlike matrices, data frames can store different classes of objects in each
  column.
- Have special attribute called: *row.names*, most of the time it is just: 1, 2,
  3, ....
- Usually created by calling *read.table()* or *read.csv()*
- Can be converted to a matrix by calling *data.matrix()* but classes will be
  unified (resulted classes will be unexpected)
  
  Example:
```{r}
x <- data.frame(foo= 1: 4, bar= c(T, T, F, F))
x
```
```{r}
nrow(x)
```
```{r}
ncol(x)
```

## Names

R objects can also have names, which is very helpful for writing readable code
and self-describing objects.

```{r}
x <- 1:3
names(x)
```
```{r}
names(x) <- c("foo", "bar", "norf")
x
```
Lists can also have names
```{r}
x <- list(a=1, b=2, c=3)
x
```
For matrices, names are basicall, the rows and columns names:
```{r}
m <- matrix(data=1:4, nrow=2, ncol=2)
dimnames(m) <- list(c('a', 'b'), c('c', 'd'))
m
```

