---
title: "Notes on R"
author: "Ahmed Eltawil"
date: "2024-10-29"
output: 
  html_document: 
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Special numbers

- *Inf*: represents infinity = 1 / 0, it can also be used as 1 / inf = 0.
         you can also have -inf.
- *NaN*: represents undefined value "Not a number"; e.g. 0 / 0; can also be
         thought as a missing value.


# Data Types

R has 5 atomic classes:

- character
- numeric ( real numbers)
- integer
- complex
- logical ( True / False )

The most basic object is a *Vector*
- A vector can only contain objects of the same class
- Exception is List, which is a vector that can contain objects of different
  classes
- Empty vector can be created by calling *vector()*

## Vectors

### Creating a vector

#### 1- Using the *c()* function
```{r, echo = TRUE}
x1 <- c(0.5, 0.6)       ## numeric
x2 <- c(TRUE, FALSE)    ## logical
x3 <- c(T, F)           ## logical
x4 <- c("a", "b", "c")  ## character
x5 <- 9:29              ## integer
x6 <- c(1+0i, 2+4i)     ## complex
```
#### 2- Using the *vector()* function
```{r}
x_v <- vector("numeric", 10)
x_v
```

### Mixing Objects
When mixing different classes in one object, all must convert into one common
class
```{r}
y <- c(1.7, "a")    ## character
y <- c(TRUE, 2)     ## numeric
y <- c("a", TRUE)   ## character
```

### Explicit Coercion
Object can be explicitly coerced from one type to another using the **as. \***
functions if available
Example:
```{r}
x <- 0:6
class(x)
```
```{r}
class(as.numeric(x))
```
```{r}
as.logical(x)
```
```{r}
as.character(x)
```

On the other side, non-sensical coercion results in **NAs**

```{r}
x <- c("a", "b", "c")
as.numeric(x)
```
```{r}
as.logical(x)
```
```{r}
as.complex(x)
```

## Lists

A special type of vector that can contain elements of different classes

```{r}
x <- list(1, "a", TRUE, 1+4i)
x
```


