---
title: "Read_From_Web"
output: html_document
date: "2024-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Webscrapping
__`Webscrapping`__: Programmatically extracting data from the HTML code of
websites.

- In some cases this is against the terms of service for the website
- Attempting to read too many pages too quickly can get your IP address
blocked

## Open connection to a website and extract the html code out of it

```{r}
con <- url("http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en")
htmlCode <- readLines(con)
close(con)
htmlCode
```

## Parsing html using XML library
```{r}
library(XML)
library(RCurl)
url <- "https://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en"
htmlContent <- getURL(url)
html <- htmlTreeParse(htmlContent, useInternalNodes = T)

xpathSApply(html, "//title", xmlValue)
```


```{r}
xpathSApply(html, "//td[@class='gsc_a_c']", xmlValue)
```

## Get from the httr package

```{r}
library(httr)
html2 <- GET(url)
content2 <- content(html2, as= "text")
parsedHtml <- htmlParse(content2, asText= TRUE)
xpathSApply(parsedHtml, "//title", xmlValue)
```

## Accessing websites with password

```{r}
pg1= GET("http://httpbin.org/basic-auth/user/passwd")
pg1
```

The error __401__ is reported because the session needs to be authenticated
first.
Authentication credentials could be sent using the _`authenticate`_ command
```{r}
pg2= GET("http://httpbin.org/basic-auth/user/passwd",
        authenticate("user", "passwd"))
pg2
```
Authentication succeed!

```{r}
names(pg2)
```

# Using handles

using handles is useful as it keeps cookies saved and it no need to
re-authenticate a page several times.

```{r}
google= handle("http://google.com")
pg1= GET(handle=google, path="/")
pg2= GET(handle=google, path="search")
```

